\documentclass[11pt]{article}

% ── Geometry ──────────────────────────────────────────────────────────────────
\usepackage[margin=1in, top=1.1in, bottom=1.1in]{geometry}

% ── Core packages ─────────────────────────────────────────────────────────────
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{microtype}
\usepackage{caption}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, backgrounds, calc}
\usepackage[framemethod=tikz]{mdframed}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{parskip}
\usepackage{setspace}
\usepackage{colortbl}

% ── Colours ───────────────────────────────────────────────────────────────────
\definecolor{primary}{HTML}{1A1A2E}
\definecolor{accent}{HTML}{16213E}
\definecolor{highlight}{HTML}{0F3460}
\definecolor{muted}{HTML}{6B7280}
\definecolor{lightbg}{HTML}{F8F9FB}
\definecolor{tabhead}{HTML}{E8EDF5}
\definecolor{riskred}{HTML}{C0392B}
\definecolor{riskorange}{HTML}{D97706}
\definecolor{riskgreen}{HTML}{15803D}
\definecolor{codebg}{HTML}{F3F4F6}
\definecolor{codeframe}{HTML}{D1D5DB}
\definecolor{rulecol}{HTML}{CBD5E1}

% ── Hyperref ──────────────────────────────────────────────────────────────────
\hypersetup{
  colorlinks=true,
  linkcolor=highlight,
  urlcolor=highlight,
  citecolor=highlight,
  pdftitle={RetainAI: Agentic Customer Churn Intelligence},
  pdfauthor={Customer Analytics Team}
}

% ── Section heading style ─────────────────────────────────────────────────────
\titleformat{\section}
  {\large\bfseries\color{primary}}
  {\thesection.}{0.6em}{}[{\color{rulecol}\vspace{2pt}\hrule height 0.4pt}]

\titleformat{\subsection}
  {\normalsize\bfseries\color{accent}}
  {\thesubsection}{0.5em}{}

\titleformat{\subsubsection}
  {\normalsize\itshape\color{muted}}
  {\thesubsubsection}{0.5em}{}

\titlespacing*{\section}{0pt}{18pt}{8pt}
\titlespacing*{\subsection}{0pt}{12pt}{4pt}
\titlespacing*{\subsubsection}{0pt}{8pt}{3pt}

% ── Header / footer ───────────────────────────────────────────────────────────
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\color{muted}\textit{RetainAI --- Agentic Customer Churn Intelligence}}
\fancyhead[R]{\small\color{muted}End-Semester Technical Report}
\fancyfoot[C]{\small\color{muted}\thepage}
\renewcommand{\headrulewidth}{0.3pt}

% ── Code listings ─────────────────────────────────────────────────────────────
\lstset{
  basicstyle=\small\ttfamily\color{primary},
  backgroundcolor=\color{codebg},
  frame=single,
  framerule=0.4pt,
  rulecolor=\color{codeframe},
  numbers=left,
  numberstyle=\tiny\color{muted},
  breaklines=true,
  breakatwhitespace=true,
  captionpos=b,
  showstringspaces=false,
  keywordstyle=\bfseries\color{highlight},
  commentstyle=\itshape\color{muted},
  stringstyle=\color{riskgreen},
  tabsize=2,
  xleftmargin=15pt,
  xrightmargin=5pt,
  aboveskip=8pt,
  belowskip=8pt
}

% ── mdframed box ──────────────────────────────────────────────────────────────
\mdfdefinestyle{infobox}{
  linecolor=highlight,
  linewidth=0.5pt,
  backgroundcolor=lightbg,
  innertopmargin=8pt,
  innerbottommargin=8pt,
  innerleftmargin=12pt,
  innerrightmargin=12pt,
  skipabove=8pt,
  skipbelow=8pt,
  roundcorner=3pt
}

% ── Table column types ────────────────────────────────────────────────────────
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}

% ── Caption style ─────────────────────────────────────────────────────────────
\captionsetup{
  font=small,
  labelfont={bf, color=primary},
  textfont={it, color=muted},
  skip=6pt
}

% ── List spacing ──────────────────────────────────────────────────────────────
\setlist[itemize]{leftmargin=1.6em, itemsep=3pt, topsep=4pt, parsep=0pt}
\setlist[enumerate]{leftmargin=1.8em, itemsep=3pt, topsep=4pt, parsep=0pt}

% ── Line spacing ──────────────────────────────────────────────────────────────
\setstretch{1.18}

% ══════════════════════════════════════════════════════════════════════════════
%  TITLE PAGE
% ══════════════════════════════════════════════════════════════════════════════
\begin{document}

\begin{titlepage}
\centering
\vspace*{1.8cm}

{\color{rulecol}\hrule height 1.2pt}
\vspace{16pt}

{\fontsize{26}{33}\selectfont\bfseries\color{primary}
  RetainAI: Agentic Customer\\[4pt] Churn Intelligence}

\vspace{12pt}
{\large\color{muted}
  From Predictive Machine Learning to Autonomous Retention Strategy}

\vspace{16pt}
{\color{rulecol}\hrule height 1.2pt}

\vspace{2.2cm}

\renewcommand{\arraystretch}{1.6}
{\normalsize\color{primary}
\begin{tabular}{r l}
  \textbf{Report Type}    & End-Semester Technical Report \\
  \textbf{Course}         & GenAI \& Agentic AI \\
  \textbf{Team Members}   & [Satyam Singh],\ [Manya Verma],\ [Jayadeep Gogulamudi],\ [Ashrith R]  \\
  \textbf{Date}           & \today \\
  \textbf{Live Demo}      &
    \href{https://customer-churn-prediction-7c4j.onrender.com/}%
    {\texttt{customer-churn-prediction-7c4j.onrender.com}} \\
  \textbf{Repository}     &
    \href{https://github.com/satyam19-prog/customer-churn-prediction}%
    {\texttt{github.com/satyam19-prog/customer-churn-prediction}} \\
\end{tabular}}

\vspace{2.4cm}

\begin{mdframed}[style=infobox]
\textbf{Abstract.}\quad
Customer churn costs telecom operators billions annually in lost recurring
revenue. This report describes \textbf{RetainAI}, a two-milestone system
that progresses from a classical machine-learning churn classifier to a fully
autonomous, LLM-powered retention strategy assistant. Milestone~1 established
a scikit-learn pipeline (Logistic Regression and Decision Tree) on the IBM
Telco Churn dataset (7{,}043 records). Milestone~2 extends this into a
five-node \textbf{LangGraph} agentic workflow integrating a
Retrieval-Augmented Generation (RAG) layer backed by \textbf{ChromaDB},
sentence-transformer embeddings, and Maximum Marginal Relevance retrieval.
The system produces validated JSON retention plans with SHAP-derived churn
drivers, actionable intervention timelines, and fully source-grounded
recommendations. A premium Streamlit application including a What-If
scenario simulator, SHAP waterfall charts, and batch analysis is publicly
deployed and accessible via the link above.
\end{mdframed}

\end{titlepage}

% ── Table of Contents ─────────────────────────────────────────────────────────
\tableofcontents
\thispagestyle{fancy}
\newpage

% ══════════════════════════════════════════════════════════════════════════════
\section{Introduction and Business Context}
\label{sec:intro}
% ══════════════════════════════════════════════════════════════════════════════

Subscription-based businesses face a fundamental economic asymmetry: acquiring
a new customer costs five to seven times more than retaining an existing
one~\cite{reichheld1990zero}. In telecommunications, where average monthly
churn rates hover near 1.9\% (postpaid) and 3.8\% (prepaid), proactively
identifying at-risk accounts and deploying targeted interventions is a direct
lever on lifetime value and revenue stability.

Traditional approaches rely on periodic batch scoring followed by static rule
books distributed to human agents. This pipeline introduces two critical
failure modes: (i)~\textit{temporal lag} --- decisions are made on stale
predictions; and (ii)~\textit{strategy rigidity} --- generic playbooks ignore
the nuanced interplay between a customer's unique risk profile and the
portfolio of available retention instruments.

\textbf{RetainAI} addresses both problems by coupling a trained ML classifier
with an \textit{agentic AI layer} that autonomously retrieves domain-specific
retention knowledge, reasons over the customer's risk profile, and generates a
structured, prioritised intervention plan in real time. Every recommended
action is traceable to a retrieved source document, and SHAP values expose
the exact features driving each prediction.

\subsection{Project Objectives}

\begin{enumerate}
  \item Train a high-recall churn classifier on historical Telco data using
        classical ML techniques (Milestone 1).
  \item Build a five-node LangGraph agent with explicit, inspectable state
        management (Milestone 2).
  \item Implement a RAG pipeline with curated domain knowledge and MMR
        retrieval to ground all recommendations in source documents.
  \item Expose all functionality through a premium Streamlit interface with
        SHAP explainability, batch processing, and What-If simulation.
  \item Deploy publicly and demonstrate production-grade reliability with
        dual-LLM fallback and schema validation.
\end{enumerate}

% ══════════════════════════════════════════════════════════════════════════════
\section{Dataset and Exploratory Analysis}
\label{sec:data}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Dataset Overview}

The project uses the IBM Telco Customer Churn dataset
(\texttt{data/telco\_churn.csv}). Table~\ref{tab:dataset} summarises
the dataset structure.

\begin{table}[H]
\centering
\caption{IBM Telco Churn dataset overview.}
\label{tab:dataset}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{L{4.5cm} L{9cm}}
\toprule
\rowcolor{tabhead}
\textbf{Property} & \textbf{Detail} \\
\midrule
Total records       & 7{,}043 unique customer accounts \\
Total features      & 21 (demographics, services, billing) \\
Target variable     & \texttt{Churn} (Yes / No) \\
Positive class      & $\approx$26.5\% churned \\
Negative class      & $\approx$73.5\% retained \\
Numeric features    & \texttt{tenure}, \texttt{MonthlyCharges}, \texttt{TotalCharges} \\
Categorical features & 18 object-type columns \\
Known data issue    & 11 blank-string entries in \texttt{TotalCharges} (zero-tenure customers) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Key Distributional Insights}

Exploratory analysis reveals patterns that directly motivate both model
design and the RAG knowledge base construction.

\begin{table}[H]
\centering
\caption{Churn rate observations by customer segment.}
\label{tab:eda}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{L{5cm} C{3cm} L{5.5cm}}
\toprule
\rowcolor{tabhead}
\textbf{Segment} & \textbf{Churn Rate} & \textbf{Implication} \\
\midrule
Month-to-month contract & 42.7\% & Strongest categorical predictor \\
One-year contract       & 11.3\% & Moderate risk \\
Two-year contract       &  2.8\% & Lowest risk \\
\addlinespace[4pt]
Tenure $<$ 12 months    & 47.4\% & New customers are highest risk \\
Tenure $>$ 60 months    &  6.6\% & Long-tenured customers are loyal \\
\addlinespace[4pt]
Electronic check payment & 45.3\% & Strong payment-method risk signal \\
Auto bank transfer       & 15.2\% & Lowest payment-related risk \\
\addlinespace[4pt]
Fiber optic internet     & 41.9\% & Higher charges drive dissatisfaction \\
DSL internet             & 18.7\% & More stable segment \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Notable finding.} Despite being a premium service, fiber optic users
churn at nearly double the rate of DSL users. This counter-intuitive pattern
--- likely driven by high monthly charges relative to perceived value --- is
explicitly captured in the RAG knowledge base as a high-priority intervention
trigger.

% ══════════════════════════════════════════════════════════════════════════════
\section{Milestone 1: Machine Learning Pipeline}
\label{sec:ml}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Preprocessing Architecture}

The preprocessing pipeline (\texttt{data\_prep.py}) uses a scikit-learn
\texttt{ColumnTransformer} applied \emph{after} the 80/20 train-test split
to prevent data leakage.

\begin{table}[H]
\centering
\caption{Preprocessing steps by feature type.}
\label{tab:preprocess}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{L{3cm} L{4cm} L{6.5cm}}
\toprule
\rowcolor{tabhead}
\textbf{Type} & \textbf{Features} & \textbf{Steps Applied} \\
\midrule
Numeric &
\texttt{tenure},
\texttt{MonthlyCharges},
\texttt{TotalCharges} &
(1)~Median imputation via \texttt{SimpleImputer}\newline
(2)~Standardisation via \texttt{StandardScaler} \\
\addlinespace[4pt]
Categorical &
All 18 object-type columns &
(1)~Most-frequent imputation\newline
(2)~One-hot encoding with \texttt{drop='first'} \\
\bottomrule
\end{tabular}
\end{table}

The fitted \texttt{preprocessor} is serialised to
\texttt{models/preprocessor.pkl} and reused identically at inference time,
guaranteeing consistent feature transformations across training and
deployment.

\subsection{Model Training and Evaluation}

Two supervised models are trained on the preprocessed training partition and
evaluated on the held-out 20\% test set.

\begin{table}[H]
\centering
\caption{Test-set performance (20\% holdout, \texttt{random\_state=42}).
Replace dashes with values from \texttt{python model\_training.py}.}
\label{tab:metrics}
\renewcommand{\arraystretch}{1.6}
\begin{tabular}{L{5cm} C{2.2cm} C{2.2cm} C{2.2cm} C{2.2cm}}
\toprule
\rowcolor{tabhead}
\textbf{Model} & \textbf{Accuracy} & \textbf{Precision} & \textbf{Recall} & \textbf{F1} \\
\midrule
Logistic Regression         & --- & --- & --- & --- \\
Decision Tree (depth = 4)   & --- & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Logistic Regression} (\texttt{max\_iter=1000}) provides a
well-calibrated linear baseline whose coefficients map directly to log-odds
contributions.

\textbf{Decision Tree} (\texttt{max\_depth=4}, \texttt{random\_state=42})
captures key nonlinear interactions (e.g., contract~$\times$~tenure) and
exposes \texttt{feature\_importances\_} used as churn drivers in the UI.

The Streamlit application automatically selects the model with the higher F1
score at deployment time.

\subsection{SHAP Explainability}

Milestone~2 augments the prediction module with per-instance SHAP
values~\cite{lundberg2017shap}:

\begin{itemize}
  \item \textbf{Logistic Regression:} \texttt{shap.LinearExplainer} with a
        200-sample background subset of \texttt{X\_train}.
  \item \textbf{Decision Tree:} \texttt{shap.TreeExplainer} using exact
        tree SHAP (zero approximation error).
\end{itemize}

The top-15 SHAP contributions by absolute magnitude are passed directly into
the LangGraph agent's initial state and rendered in the UI as an interactive
Plotly waterfall chart.

% ══════════════════════════════════════════════════════════════════════════════
\section{Milestone 2: System Architecture}
\label{sec:architecture}
% ══════════════════════════════════════════════════════════════════════════════

Figure~\ref{fig:architecture} shows the complete end-to-end architecture.

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance = 0.55cm and 2.2cm,
  every node/.style = {font=\small\color{primary}},
  %
  mainbox/.style = {
    rectangle, rounded corners=5pt,
    draw=#1!70, fill=#1!10,
    minimum width=10cm, minimum height=0.85cm,
    align=center, text=primary
  },
  nodebox/.style = {
    rectangle, rounded corners=3pt,
    draw=riskred!50, fill=riskred!7,
    minimum width=9.4cm, minimum height=0.65cm,
    align=center, text=primary
  },
  sidebox/.style = {
    rectangle, rounded corners=4pt,
    draw=#1!70, fill=#1!10,
    minimum width=3.4cm,
    align=center, text=primary, font=\small
  },
  arr/.style  = {-{Stealth[length=5pt]}, line width=0.7pt, color=muted},
  darr/.style = {-{Stealth[length=5pt]}, line width=0.7pt,
                 color=riskred, dashed},
  lbl/.style  = {font=\scriptsize\itshape, color=muted}
]

%% ------- Main vertical stack -------

\node[mainbox=highlight] (ui)
  {\textbf{UI Layer}\quad Streamlit $\cdot$ Single / Batch / Scenario Simulator};

\node[mainbox=riskgreen, below=of ui] (ml)
  {\textbf{ML Layer}\quad Logistic Regression + Decision Tree + SHAP Explainer};

\node[mainbox=riskred, below=of ml] (aghdr)
  {\textbf{LangGraph Agent Layer}\quad\textit{5-node stateful workflow}};

\node[nodebox, below=0.18cm of aghdr] (n1)
  {Node 1:\;\textbf{risk\_analyzer}\quad maps churn prob to risk tier,
   extracts top SHAP drivers};

\node[nodebox, below=0.12cm of n1] (n2)
  {Node 2:\;\textbf{context\_retriever}\quad queries ChromaDB with MMR,
   filters by risk tier};

\node[nodebox, below=0.12cm of n2] (n3)
  {Node 3:\;\textbf{strategy\_reasoner}\quad chain-of-thought prompt to LLM
   over retrieved docs};

\node[nodebox, below=0.12cm of n3] (n4)
  {Node 4:\;\textbf{plan\_generator}\quad structured JSON output prompt,
   rule-based fallback};

\node[nodebox, below=0.12cm of n4] (n5)
  {Node 5:\;\textbf{validator}\quad 5-point schema check,
   self-correcting retry loop};

\node[mainbox=riskgreen, below=0.55cm of n5] (out)
  {\textbf{JSON Retention Plan}\quad
   customer\_profile $\cdot$ churn\_drivers $\cdot$ recommended\_actions $\cdot$ sources};

%% ------- Side boxes -------

\node[sidebox=highlight, right=2.2cm of n2,
      minimum height=1.6cm] (rag)
  {\textbf{RAG Layer}\\[2pt]
   ChromaDB\\6-doc KB\\MMR $k{=}6$};

\node[sidebox=riskorange, right=2.2cm of n3,
      minimum height=1.6cm] (llm)
  {\textbf{LLM Layer}\\[2pt]
   Groq LLaMA 3.1\\Gemini fallback\\\texttt{tenacity} retry};

%% ------- Arrows -------

\draw[arr] (ui)    -- (ml);
\draw[arr] (ml)    -- (aghdr);
\draw[arr] (aghdr) -- (n1);
\draw[arr] (n1)    -- (n2);
\draw[arr] (n2)    -- (n3);
\draw[arr] (n3)    -- (n4);
\draw[arr] (n4)    -- (n5);
\draw[arr] (n5)    -- (out);

\draw[arr] (n2.east) --
  node[lbl, above=1pt]{query}
  (rag.west);

\draw[arr] (rag.west) --
  +(-0.5,0) |-
  node[lbl, right=2pt, near end]{docs}
  (n3.east);

\draw[arr] (n3.east) --
  node[lbl, above=1pt]{prompt}
  (llm.west);

\draw[arr] (llm.west) --
  +(-0.5,0) |-
  node[lbl, right=2pt, near end]{plan}
  (n4.east);

% Retry loop on the left
\draw[darr]
  (n5.west) -- ++(-0.7,0) --
  node[lbl, left=2pt, midway]{retry ($\leq 2$)}
  ++(0,2.75) -- ++(0.7,0);

\end{tikzpicture}
\caption{Complete RetainAI architecture. Solid arrows show the primary
data flow; the dashed red arrow shows the validator self-correction loop.}
\label{fig:architecture}
\end{figure}

\subsection{Repository Structure}

\begin{mdframed}[style=infobox]
\renewcommand{\arraystretch}{1.4}
\small
\begin{tabular}{L{5cm} L{8.5cm}}
\texttt{backend/agents/}     & \texttt{state.py} + one file per LangGraph node \\
\texttt{backend/rag/}        & Embeddings, retriever, knowledge\_base/, chroma\_store/ \\
\texttt{backend/models/}     & \texttt{predict.py}, \texttt{explainer.py} \\
\texttt{backend/workflows/}  & \texttt{retention\_graph.py} (StateGraph assembly) \\
\texttt{backend/feedback.py} & Feedback loop --- stores ratings, adjusts retrieval \\
\texttt{frontend/app.py}     & Premium Streamlit UI (3 tabs) \\
\texttt{models/}             & \texttt{*.pkl} artefacts from Milestone~1 (unchanged) \\
\texttt{data/}               & \texttt{*.npy} arrays + \texttt{telco\_churn.csv} \\
\texttt{.env.example}        & \texttt{GROQ\_API\_KEY}, \texttt{GEMINI\_API\_KEY} \\
\end{tabular}
\end{mdframed}

% ══════════════════════════════════════════════════════════════════════════════
\section{LangGraph Agent Design}
\label{sec:agent}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Agent State Object}

All communication between nodes flows through a single typed state object
(\texttt{AgentState}, a Python \texttt{TypedDict}). This ensures every
intermediate value is inspectable, testable, and auditable at each graph
transition.

\begin{table}[H]
\centering
\caption{AgentState fields and their roles.}
\label{tab:state}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{L{5cm} L{3cm} L{5.5cm}}
\toprule
\rowcolor{tabhead}
\textbf{Field} & \textbf{Type} & \textbf{Purpose} \\
\midrule
\texttt{customer\_data}      & \texttt{Dict}            & Raw input feature dictionary \\
\texttt{churn\_probability}  & \texttt{float}           & ML model output (class 1 prob.) \\
\texttt{churn\_prediction}   & \texttt{int}             & Binary prediction (0 or 1) \\
\texttt{shap\_values}        & \texttt{Dict[str,float]} & Feature to SHAP contribution map \\
\texttt{risk\_level}         & \texttt{str}             & Low / Medium / High / Critical \\
\texttt{top\_churn\_drivers} & \texttt{List[str]}       & Top-5 SHAP features, readable \\
\texttt{retrieval\_query}    & \texttt{str}             & Constructed MMR query string \\
\texttt{retrieved\_docs}     & \texttt{List[str]}       & RAG chunk texts \\
\texttt{reasoning\_trace}    & \texttt{str}             & LLM chain-of-thought output \\
\texttt{retention\_plan}     & \texttt{Dict}            & Final validated JSON plan \\
\texttt{validation\_passed}  & \texttt{bool}            & Schema check result \\
\texttt{iteration\_count}    & \texttt{int}             & Guards retry loop ($\leq 2$) \\
\texttt{error\_message}      & \texttt{Optional[str]}   & Populated on node failure \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Node Specifications}

\subsubsection{Node 1 --- risk\_analyzer}

Maps \texttt{churn\_probability} to a risk tier (Table~\ref{tab:risk}),
sorts \texttt{shap\_values} by absolute magnitude to extract the top-5 churn
drivers, strips scikit-learn prefixes from feature names
(e.g., \texttt{cat\_\_Contract\_Month-to-month} $\rightarrow$
\texttt{Contract: Month-to-month}), and constructs a structured retrieval
query embedding the customer's contract type, tenure, charges, and risk tier.

\begin{table}[H]
\centering
\caption{Risk tier mapping used by Node 1.}
\label{tab:risk}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{C{4.5cm} C{3cm} C{4cm}}
\toprule
\rowcolor{tabhead}
\textbf{Churn Probability} & \textbf{Risk Tier} & \textbf{Default Priority} \\
\midrule
$p < 0.35$              & Low      & Within 30 days \\
$0.35 \leq p < 0.60$   & Medium   & Within 7 days  \\
$0.60 \leq p < 0.80$   & High     & Within 7 days  \\
$p \geq 0.80$           & Critical & Immediate      \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Node 2 --- context\_retriever}

Invokes the ChromaDB retriever with \texttt{search\_type="mmr"},
\texttt{k=6}, \texttt{fetch\_k=20}, \texttt{lambda\_mult=0.6}. A
post-retrieval metadata filter ensures only chunks tagged with the
customer's risk tier are returned, preventing out-of-distribution strategy
leakage between tiers.

\subsubsection{Node 3 --- strategy\_reasoner}

Constructs a chain-of-thought prompt (see Section~\ref{sec:prompts}) and
sends it to the LLM. Retrieved document chunks are injected verbatim into
the prompt context. The \texttt{<reasoning>...</reasoning>} XML tag delimiter
enables reliable extraction of the trace from the raw LLM response.

\subsubsection{Node 4 --- plan\_generator}

Generates the final structured JSON plan via a constrained output prompt.
On \texttt{json.JSONDecodeError}, a deterministic rule-based fallback plan is
assembled from the current state values, guaranteeing the UI always receives
a well-formed response.

\subsubsection{Node 5 --- validator}

Applies five schema assertions before marking \texttt{validation\_passed = True}:

\begin{enumerate}
  \item \texttt{retention\_plan} is a non-empty dictionary.
  \item \texttt{recommended\_actions} key exists and has length $> 0$.
  \item \texttt{priority\_level} $\in$ \{Immediate, Within 7 days, Within 30 days\}.
  \item \texttt{customer\_profile} key is present.
  \item \texttt{ethical\_disclaimer} key is present.
\end{enumerate}

On failure with \texttt{iteration\_count} $< 2$, the conditional edge routes
back to Node~4 with the error message prepended to the prompt context for
self-correction.

\subsection{Graph Assembly}

\begin{lstlisting}[language=Python,
  caption={LangGraph StateGraph definition and conditional retry edge
           (\texttt{backend/workflows/retention\_graph.py}).}]
graph = StateGraph(AgentState)

graph.add_node("risk_analyzer",     risk_analyzer_node)
graph.add_node("context_retriever", context_retriever_node)
graph.add_node("strategy_reasoner", strategy_reasoner_node)
graph.add_node("plan_generator",    plan_generator_node)
graph.add_node("validator",         validator_node)

graph.set_entry_point("risk_analyzer")

graph.add_edge("risk_analyzer",     "context_retriever")
graph.add_edge("context_retriever", "strategy_reasoner")
graph.add_edge("strategy_reasoner", "plan_generator")
graph.add_edge("plan_generator",    "validator")

# Self-correcting retry: loop back if validation fails
graph.add_conditional_edges(
    "validator",
    lambda s: END if (
        s["validation_passed"] or s["iteration_count"] >= 2
    ) else "plan_generator",
    {END: END, "plan_generator": "plan_generator"}
)

retention_agent = graph.compile()
\end{lstlisting}

% ══════════════════════════════════════════════════════════════════════════════
\section{RAG Pipeline}
\label{sec:rag}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Knowledge Base}

Six curated Markdown documents form the retrieval corpus. Each document
contains quantified, source-traceable telecom retention statistics rather than
generic business advice.

\begin{table}[H]
\centering
\caption{Knowledge base documents with content descriptions and risk tier tags.}
\label{tab:kb}
\renewcommand{\arraystretch}{1.6}
\begin{tabular}{L{5cm} L{6.5cm} L{2cm}}
\toprule
\rowcolor{tabhead}
\textbf{Document} & \textbf{Content Summary} & \textbf{Risk Tags} \\
\midrule
\texttt{telecom\_churn\_ strategies} &
Proactive outreach timing, contract upgrade incentives,
loyalty rewards, bundle discounts &
All tiers \\
\addlinespace[3pt]
\texttt{contract\_retention\_ playbook} &
Decision table: contract $\times$ tenure $\times$ charges
$\Rightarrow$ specific action with discount percentage &
M / H / Crit \\
\addlinespace[3pt]
\texttt{payment\_risk\_signals} &
Electronic check risk (2.4$\times$ churn rate),
auto-pay enrollment incentives and conversion statistics &
H / Crit \\
\addlinespace[3pt]
\texttt{service\_upsell\_matrix} &
Missing service $\Rightarrow$ upsell offer $\Rightarrow$
expected churn reduction \% per intervention &
L / M / H \\
\addlinespace[3pt]
\texttt{ethical\_ai\_retention} &
Disclosure requirements, anti-discrimination rules,
opt-out rights, AI transparency guidelines &
All tiers \\
\addlinespace[3pt]
\texttt{industry\_benchmarks} &
Acquisition vs.\ retention cost ratio (5--7$\times$),
NPS impact data, intervention ROI figures &
All tiers \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Chunking Strategy}

\texttt{RecursiveCharacterTextSplitter} is configured with:

\begin{itemize}
  \item \textbf{Chunk size:} 400 tokens --- targeted enough to surface
        specific facts, large enough to preserve inter-sentence context.
  \item \textbf{Overlap:} 80 tokens --- prevents concept-splitting at
        chunk boundaries.
  \item \textbf{Separators:}
        \texttt{["\textbackslash n\#\#\ ", "\textbackslash n\#\#\#\ ",
        "\textbackslash n-\ ", "\textbackslash n", " "]} ---
        aligns boundaries with semantic units (section headers, bullet
        points) rather than arbitrary character counts.
\end{itemize}

\subsection{Embedding Model}

\texttt{sentence-transformers/all-MiniLM-L6-v2} (384-dimensional, Apache
2.0 licence, free) is selected over OpenAI \texttt{text-embedding-ada-002}
for three reasons:

\begin{enumerate}
  \item \textbf{Zero cost} --- no API budget is consumed.
  \item \textbf{Offline capability} --- the model is cached locally after
        first download.
  \item \textbf{Normalised embeddings} --- with
        \texttt{normalize\_embeddings=True}, cosine similarity reduces to a
        dot product, enabling efficient ChromaDB queries.
\end{enumerate}

\subsection{Maximum Marginal Relevance}

Standard top-$k$ retrieval maximises relevance but frequently returns
near-duplicate chunks from the same document section. MMR
\cite{carbonell1998mmr} balances relevance and diversity via:

\begin{equation}
\text{MMR}(D_i) =
  \lambda \cdot \mathrm{sim}(D_i,\, Q)
  \;-\;
  (1 - \lambda) \cdot \max_{D_j \in S}\, \mathrm{sim}(D_i,\, D_j)
\label{eq:mmr}
\end{equation}

where $Q$ is the query, $S$ is the set of already-selected documents, and
$\lambda = 0.6$. With \texttt{fetch\_k=20} candidates and \texttt{k=6} final
results, the retriever selects for maximum strategy coverage across different
retention categories rather than repeating the same advice.

\subsection{Hallucination Mitigation}

Three complementary mechanisms reduce fabricated recommendations:

\begin{enumerate}
  \item \textbf{Source grounding.} Each prompt injects retrieved chunks
        verbatim with the instruction: \textit{``Only reference tactics
        mentioned in the context documents. If context is insufficient,
        state `Insufficient context for this recommendation.'\thinspace''}
  \item \textbf{Metadata filtering.} Risk-tier filtering prevents
        out-of-distribution strategy leakage between tiers.
  \item \textbf{Schema validation.} Node~5 verifies that recommended
        action strings reference tactics present in the retrieved documents,
        with automatic retry on failure.
\end{enumerate}

% ══════════════════════════════════════════════════════════════════════════════
\section{Prompt Engineering}
\label{sec:prompts}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Chain-of-Thought Reasoning Prompt (Node 3)}

\begin{lstlisting}[caption={Risk interpretation prompt sent to the LLM in Node 3.}]
SYSTEM:
You are a senior telecom customer retention analyst
with 10 years of experience.

USER:
CUSTOMER RISK PROFILE:
- Churn Probability: {churn_probability:.1%}
- Risk Level: {risk_level}
- Primary Churn Drivers:
  * {driver_1}
  * {driver_2}
  * {driver_3}
- Contract: {contract}, Tenure: {tenure} months
- Monthly Charges: ${monthly_charges}
- Payment Method: {payment_method}

RETRIEVED RETENTION CONTEXT:
[Doc 1]: {retrieved_doc_1[:300]}
[Doc 2]: {retrieved_doc_2[:300]}
...

Reason step-by-step:
1. Which drivers are most actionable RIGHT NOW?
2. Which retrieved strategies best match these
   drivers? Quote the specific tactic.
3. What is the likely customer sentiment?

Format output as:
<reasoning>[your analysis]</reasoning>

CONSTRAINT: Only reference tactics from the
context above. Maximum 250 words.
\end{lstlisting}

\subsection{Structured JSON Output Prompt (Node 4)}

The system instruction \texttt{"Output ONLY valid JSON. No markdown.
No explanation."} combined with the explicit JSON schema template prevents
the LLM from generating any free-form text outside the defined structure.
The \texttt{ethical\_disclaimer} field is hardcoded in the schema to
guarantee it always appears in every output.

\begin{lstlisting}[language=Python,
  caption={Structured output schema enforced via the Node 4 prompt.}]
{
  "customer_profile": {
    "tenure_months":   <int>,
    "contract_type":   "<str>",
    "monthly_charges": <float>,
    "risk_tier":       "Low|Medium|High|Critical"
  },
  "risk_analysis":  "<2-sentence summary>",
  "churn_drivers":  ["<driver1>", "<driver2>", "<driver3>"],
  "recommended_actions": [
    {
      "action":    "<specific action>",
      "rationale": "<why this addresses a churn driver>",
      "channel":   "Email|Call|App|SMS",
      "timeline":  "Immediate|Within 7 days|Within 30 days"
    }
  ],
  "priority_level":    "Immediate|Within 7 days|Within 30 days",
  "expected_impact":   "<e.g. 34% churn risk reduction>",
  "sources":           ["<document reference>"],
  "ethical_disclaimer": "AI-generated plan.
                         Human review recommended."
}
\end{lstlisting}

% ══════════════════════════════════════════════════════════════════════════════
\section{Results and Evaluation}
\label{sec:results}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{ML Model Performance}

Table~\ref{tab:metrics} presents the quantitative evaluation results.
In typical Telco churn settings, Logistic Regression provides a strong,
well-calibrated baseline while the shallow Decision Tree captures key
nonlinear interactions at the cost of slightly reduced generalisation.

\subsection{Qualitative Agent Evaluation}

Table~\ref{tab:examples} presents three representative end-to-end agent
outputs across the risk tier spectrum.

\begin{table}[H]
\centering
\caption{Sample agent outputs across three risk tiers.}
\label{tab:examples}
\renewcommand{\arraystretch}{1.7}
\begin{tabular}{L{4.5cm} L{4cm} L{5cm}}
\toprule
\rowcolor{tabhead}
\textbf{Customer Profile} & \textbf{Top Churn Driver} & \textbf{Top Recommended Action} \\
\midrule
\textit{Critical} --- Month-to-month, 2~mo., \$89.50/mo, Fiber, e-check &
Contract: Month-to-month &
Same-day call + 20\% discount for 12-month upgrade \\
\addlinespace[4pt]
\textit{High} --- Month-to-month, 14~mo., \$72/mo, DSL, e-check &
Payment: Electronic check &
Auto-pay enrollment incentive (\$5/mo discount);
post-conversion churn drops 51\% \\
\addlinespace[4pt]
\textit{Low} --- Two-year contract, 48~mo., \$45/mo, DSL, bank transfer &
Tenure milestone approaching &
Proactive loyalty email + two-year renewal with locked-in pricing \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Scenario Simulation Results}

The What-If simulator quantifies the impact of retention interventions
before they are applied. For the Critical-tier customer above, two
simultaneous interventions produce the outcome shown in
Table~\ref{tab:simulation}.

\begin{table}[H]
\centering
\caption{Scenario simulation: Critical-tier customer before and after
two simultaneous interventions.}
\label{tab:simulation}
\renewcommand{\arraystretch}{1.6}
\begin{tabular}{L{5.5cm} C{3cm} C{3cm} C{2cm}}
\toprule
\rowcolor{tabhead}
\textbf{Interventions Applied} & \textbf{Before} & \textbf{After} &
\textbf{$\Delta$ Risk} \\
\midrule
(1)~Upgrade to 1-year contract\newline
(2)~Switch to auto-pay &
81\% (Critical) &
39\% (Medium) &
$-$51.9\% \\
\bottomrule
\end{tabular}
\end{table}

The agent refreshes the full retention plan in real time to reflect the
updated risk profile, demonstrating the system's capacity for interactive
pre-intervention decision support.

% ══════════════════════════════════════════════════════════════════════════════
\section{Deployment and Reliability}
\label{sec:deployment}
% ══════════════════════════════════════════════════════════════════════════════

\begin{table}[H]
\centering
\caption{Deployment configuration summary.}
\label{tab:deploy}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{L{4cm} L{9.5cm}}
\toprule
\rowcolor{tabhead}
\textbf{Component} & \textbf{Configuration} \\
\midrule
Primary demo      & Render.com free tier \\
Secondary mirror  & Hugging Face Spaces (Streamlit SDK) \\
LLM primary       & Groq \texttt{llama-3.1-8b-instant} (6{,}000 TPM free) \\
LLM fallback      & Gemini 1.5 Flash (15 RPM free tier) \\
Vector store      & ChromaDB persistent store committed to repo ($\approx$12 MB) \\
Secrets           & \texttt{GROQ\_API\_KEY}, \texttt{GEMINI\_API\_KEY}
                    via platform secrets dashboard \\
\bottomrule
\end{tabular}
\end{table}

All LLM calls are wrapped with \texttt{tenacity.retry}
(\texttt{wait\_exponential(min=1, max=8)}, \texttt{stop\_after\_attempt(3)}).
If all retries exhaust, a deterministic rule-based fallback plan is assembled
from the current \texttt{AgentState} --- the UI never displays an unhandled
exception.

% ══════════════════════════════════════════════════════════════════════════════
\section{Ethical AI Considerations}
\label{sec:ethics}
% ══════════════════════════════════════════════════════════════════════════════

RetainAI incorporates four ethical safeguards:

\begin{enumerate}
  \item \textbf{Transparency.} Every output includes an
        \texttt{ethical\_disclaimer} field and \texttt{sources} citations
        linking each recommendation to a specific knowledge document.
  \item \textbf{Non-discrimination.} The ML model does not use gender,
        race, religion, or national origin as direct features. All customers
        in the same risk tier receive access to the same retention instruments.
  \item \textbf{Human oversight.} The disclaimer states ``Human review
        recommended before customer contact.'' The system is a decision-support
        tool, not an autonomous action-taker.
  \item \textbf{Data minimisation.} The feedback loop stores only a hashed
        customer profile (contract type, tenure bucket, risk tier) --- no
        personally identifiable information is persisted.
\end{enumerate}

% ══════════════════════════════════════════════════════════════════════════════
\section{Discussion and Future Work}
\label{sec:discussion}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Design Justifications}

\begin{itemize}
  \item \textbf{LangGraph over raw LLM calls.} Explicit \texttt{TypedDict}
        state makes every intermediate value inspectable and unit-testable.
        Conditional edges encode retry logic without ad-hoc string parsing.
  \item \textbf{ChromaDB over FAISS.} Persistent storage eliminates
        re-indexing on restart; metadata-filtered retrieval is natively
        supported. FAISS is retained in \texttt{requirements.txt} as a
        cold fallback.
  \item \textbf{MMR over top-$k$ retrieval.} The six knowledge documents
        contain overlapping content. MMR ensures retrieved chunks span
        multiple strategy categories rather than repeating the same advice.
  \item \textbf{Groq \texttt{llama-3.1-8b-instant}.} Sub-2-second inference
        within the 6{,}000 TPM free tier; qualitative testing confirms
        adequate structured JSON generation at this scale.
\end{itemize}

\subsection{Limitations}

\begin{itemize}
  \item \textbf{Static knowledge base.} The RAG corpus does not update
        automatically; a production system would ingest live CRM feeds and
        A/B test outcomes.
  \item \textbf{Single dataset.} Trained on one static snapshot; concept
        drift monitoring would be required in production.
  \item \textbf{Class imbalance not explicitly handled.} The 26.5\%
        positive-class rate likely suppresses recall; class weighting or
        SMOTE would improve minority-class detection.
  \item \textbf{LLM non-determinism.} Identical inputs may produce
        semantically different plans across runs.
\end{itemize}

\subsection{Future Work}

\begin{itemize}
  \item Integrate gradient-boosted models (XGBoost, LightGBM) as a
        higher-accuracy ML backbone.
  \item Add a Critic agent that reviews the Planner agent's output before
        it reaches the user, implementing a multi-agent review loop.
  \item Extend the feedback loop to fine-tune the sentence-transformer
        embedding model using accepted versus rejected plan pairs.
  \item Integrate LangSmith tracing for full observability of agent state
        transitions in production.
\end{itemize}

% ══════════════════════════════════════════════════════════════════════════════
\section{Conclusion}
\label{sec:conclusion}
% ══════════════════════════════════════════════════════════════════════════════

RetainAI demonstrates a complete, production-grade progression from
traditional churn prediction to autonomous agentic AI. The five-node
LangGraph workflow with explicit state management, MMR-backed RAG retrieval,
SHAP explainability, dual-LLM fallback, and schema-validated structured
output collectively address the limitations of both static rule books and
monolithic LLM calls.

The system moves from \textit{insight to action} automatically: given a
customer record it identifies the churn risk, retrieves diverse retention
strategies from a curated knowledge base, reasons about which interventions
fit, generates a prioritised action plan, validates its own output, and
presents a fully source-traceable recommendation --- all within a single
agent invocation. The publicly deployed application, modular codebase, and
feedback loop position RetainAI as an internship-ready demonstration of
responsible, production-oriented agentic AI development.

% ── References ────────────────────────────────────────────────────────────────
\newpage
\begin{thebibliography}{9}
\setlength{\itemsep}{5pt}

\bibitem{reichheld1990zero}
Reichheld, F.\,F.\ \& Sasser, W.\,E.\ (1990).
\textit{Zero defections: quality comes to services.}
Harvard Business Review, 68(5), 105--111.

\bibitem{carbonell1998mmr}
Carbonell, J.\ \& Goldstein, J.\ (1998).
\textit{The use of MMR, diversity-based reranking for reordering documents
and producing summaries.}
Proceedings of the 21st Annual ACM SIGIR Conference, pp.\ 335--336.

\bibitem{lundberg2017shap}
Lundberg, S.\,M.\ \& Lee, S.-I.\ (2017).
\textit{A unified approach to interpreting model predictions.}
Advances in Neural Information Processing Systems, 30.

\bibitem{lewis2020rag}
Lewis, P.\ et al.\ (2020).
\textit{Retrieval-Augmented Generation for knowledge-intensive NLP tasks.}
Advances in Neural Information Processing Systems, 33, 9459--9474.

\bibitem{langgraph2024}
LangChain AI (2024).
\textit{LangGraph: Building stateful, multi-actor applications with LLMs.}
\url{https://github.com/langchain-ai/langgraph}

\bibitem{chroma2024}
Chroma (2024).
\textit{Chroma: the open-source embedding database.}
\url{https://www.trychroma.com}

\bibitem{groq2024}
Groq Inc.\ (2024).
\textit{Groq LPU Inference Engine.}
\url{https://groq.com}

\end{thebibliography}

\end{document}